---
title: "ghgtools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ghgtools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<br>
**ghgtools** is designed to make greenhouse gas (GHG) accounting more accessible and standardized. Tracking, measuring, and reporting GHG emissions is essential to understand risks and identify opportunities related to climate change and the transition towards a renewable energy economy. The goal of this package is to advance the practice of GHG accounting with tools rooted in uncompromising transparency, rigorous data quality, and purposeful versatility. 

## The Basics

The GHG inventory of any entity can be summarized as follows: _Assets_, such as buildings, vehicles, and equipment, engage in _Activities_, such as electricity consumption, travel, or purchasing, which each have a specific _Emission Factor_ that tells us the rate at which that activity generates GHG emissions. These three variables - assets, actives, and emission factors - compose the underlying calculation methodology. ghgtools uses a set of standardized data templates to match each record of activity data to an emission factor using information about your assets. Our methodology is built on the best practices set forth by the [GHG Protocol Corporate Standard](https://ghgprotocol.org/corporate-standard).

The following data sets are required to use ghgtools and calculate GHG emissions. When you library(ghgtools), these data sets are available.

**Emission Factor Library** - [EFLibrary](https://ghgtools.io/reference/EFLibrary.html) - The catalog of emission factors for a variety of activities. Our team consolidates emission factors from a collection of reputable sources, such as the EPA's [Emission Factor Hub](https://www.epa.gov/climateleadership/ghg-emission-factors-hub). Unless you want to add your own custom emission factors (which you certainly can, these tools are open source after all!), you don't need to worry about the emission factor library. It is pre-populated and ready to go. 

**Asset Portfolio** - [AssetPortfolio](https://ghgtools.io/reference/AssetPortfolio.html) - The list of all your assets. This likely includes buildings and vehicles. You may also designate an _Enterprise_ asset to calculate scope 3 emissions for purchasing, business travel, and other supply chain related activities. 

**Activity Data** - [ActivityData](https://ghgtools.io/reference/ActivityData.html) - The record of energy consumption and other GHG-producing activities across each of your assets. 

## Data and Templates

First, we need to load ghgtools

```{r setup}
library(ghgtools)
```

The emission factor library, `EFLibrary`, is loaded from the ghgtools package. There is a row of data for each greenhouse gas produced by the activity. For example, burning natural gas will generate CO2, CH4 and N2O. `EFLibrary` is loaded when you library ghgtools. An example of the emission factors for _Aviation Gasoline_ is provided below. Check out the [Emission Factor Library](https://ghgtools.io/articles/Emission-Factor-Library.html) article for more guidance and information. 

<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)
EFLibrary[is.na(EFLibrary)] <- ""
kbl(head(EFLibrary, 3)) %>%
  kable_material(c("striped", "hover")) %>%
  scroll_box(width = "100%", height = "100%")
```

<br>
ghgtools comes with templates for activity data and the asset portoflio. You can use `create_templates()` to write these templates to excel files in your working directory. `create_templates()` also writes the full emission factor library to an excel file in your working directory. 
<br>

```{r, include=TRUE, message=FALSE, warning=FALSE}
create_templates()
```

See below for some example data from the templates. 

**Activity Data Template** - _See_ [_ActivityData_](https://ghgtools.io/reference/ActivityData.html) _reference documentation for more information about the activity data template._
<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ActivityData[is.na(ActivityData)] <- ""
kbl(head(ActivityData, 4)) %>%
  kable_material(c("striped", "hover")) %>%
  scroll_box(width = "100%", height = "100%")
```

<br>
**Asset Portfolio Template** - _See_ [_AssetPortfolio_](https://ghgtools.io/reference/AssetPortfolio.html) _reference documentation for more information about the asset portfolio template._
<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
AssetPortfolio[is.na(AssetPortfolio)] <- ""
kbl(head(AssetPortfolio, 3)) %>%
  kable_material(c("striped", "hover")) %>%
  scroll_box(width = "100%", height = "100%")
```

<br>

## Running ghg_inventory()

The core function of ghgtools is `ghg_inventory()`. This function requires activity data, an asset portoflio, and a GWP selection.You may choose to provide an emission factor library, otherwise `ghg_inventory()` will default to the internal emission factor library `EFLibrary`

- **Activity Data** must be a data frame following the structure of the activity data template. 
- **Asset Portfolio** must be a data frame following the structure of the asset portfolio template. 
- **GWP** choose a global wamring potential to use for the GHG inventory calcualtions. Choices are “SAR”, “AR4”, “AR5”, or “AR6”. If you have no preference for GWPs, we recommend following UNFCCC guidelines, which require the use of GWP values from the IPCC’s Fifth Assessment Report (AR5). 
- **Emission Factor Library** must be a data frame following the structure of `EFLibrary`. You may have custom or unique emission factors that you wish to use in ghgtools, in which case you can load your own emission factor library into r studio. You may also append additional emission factors onto the default `EFLibrary`. 

You may prefer to use excel to overwrite the activity data and asset portfolio data templates with your own data. This is why `create_templates()` writes an excel file in your working directory. Once you add your data to the templates, you can use the `fread()` function to create global environment variables for use in ghgtools. If you decide to make any changes to the emission factor library, again use `fread()` to create a global environment variable. 
<br>

```{r, include=TRUE, message=FALSE, warning=FALSE}
library(data.table)
My_ActivityData <- fread("ActivityData.csv")
My_AssetPortfolio <- fread("AssetPortfolio.csv")
My_EFLibrary <- fread("EFLibrary.csv")
```

With the variables created above, you can run `ghg_inventory()`.

```{r, include=TRUE, message=FALSE, warning=FALSE}
GHG_rawdata <- ghg_inventory(My_ActivityData, My_AssetPortfolio, "AR5", My_EFLibrary)
```

Let's take a look at the ghg inventory that was calculated:
<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
kbl(head(GHG_rawdata, 100)) %>%
  kable_material(c("striped", "hover")) %>%
  scroll_box(width = "100%", height = "400px")
```

<br>
`ghg_inventory()` has merged our AssetPortoflio, ActivityData, and EFL to generate a GHG emissions inventory report. This data can now be used for a variety of climate disclosure needs. 

## Visualize your GHG Inventory

Using `ggplot2()` we can begin to visualize the inventory created by ghgtools.  
<br>

```{r, echo=FALSE, message = FALSE, warning=FALSE, fig.width=6, fig.height=4}
library(ggplot2)
ghg_summary <- GHG_rawdata[, .(EmissionTotal = sum(mt_co2e)), by = emission_scope]
ghg_summary[, percentage := (EmissionTotal / sum(EmissionTotal))*100]
GHG_sum_chart <- ggplot(ghg_summary, aes(x = emission_scope, y = EmissionTotal)) +
  geom_bar(stat = "identity", width = 0.8, fill = "grey", color = "black") +
  geom_text(aes(label = paste0(round(percentage), "%")),
            position = position_stack(vjust = 0.5),
            size = 4,
            fontface = "bold",
            color = "black") +
  labs(title = "GHG Emissions Total",
       x = "",
       y = "MT CO2e")
print(GHG_sum_chart)
```

<br>

```{r, echo=FALSE, message = FALSE, warning=FALSE, fig.width=6, fig.height=4}
Ecat_summary <- GHG_rawdata[, .(EmissionTotal = sum(mt_co2e)), by = service_type]
Ecat_sum_chart <- ggplot(Ecat_summary, aes(x = reorder(service_type, EmissionTotal), y = EmissionTotal)) +
  geom_bar(stat = "identity", width = 0.7, fill = "darkgreen", color = "black") +
  labs(title = "GHG Emissions by Activity Type",
       x = "",
       y = "MT CO2e") +
  coord_flip()
print(Ecat_sum_chart)
```

<br>

```{r, echo=FALSE, message = FALSE, warning=FALSE,fig.width=6, fig.height=4}
scope1and2 <- c("Scope 1", "Scope 2")
state_data <- GHG_rawdata[emission_scope %in% scope1and2, .(EmissionTotal = sum(mt_co2e)), by = state]
state_data[, EmissionTotal := round(EmissionTotal, 0)]
state_data[, mid := EmissionTotal - (0.5 * EmissionTotal)]
state_data_chart <- ggplot(state_data, aes(x = state, y = EmissionTotal)) +
  geom_bar(stat = "identity", width = 0.5, fill = "lightgrey", color = "black") +
   geom_text(aes(x = state, y = mid, label = paste0(EmissionTotal, " MT CO2e")), 
            size = 4,
            fontface = "bold", 
            color = "black") +
  labs(title = "Scope 1 & 2 Emissions by State",
       x = "",
       y = "") +
  coord_flip()
print(state_data_chart)
```

<br>

Visit [ghgtools.io](ghgtools.io) for updates and more information!
