---
title: "ghgtools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ghgtools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<br>
**ghgtools** is designed to make greenhouse gas (GHG) accounting more accessible and standardized. Tracking, measuring, and reporting GHG emissions is essential to understand risks and identify opportunities related to climate change and the transition towards a renewable energy economy. The goal of this package is to advance the practice of GHG accounting with tools rooted in uncompromising transparency, rigorous data quality, and purposeful versatility. 

## The Basics

The GHG inventory of any entity can be summarized as follows: _Assets_, such as buildings, vehicles, and equipment, engage in _Activities_, such as electricity consumption, travel, or purchasing, which each have a specific _Emission Factor_ that tells us the rate at which that activity generates GHG emissions. These three variables - assets, actives, and emission factors - compose the underlying calculation methodology. ghgtools uses a set of standardized data templates to match each record of activity data to an emission factor using information about your assets. Our methodology is built on the best practices set forth by the [GHG Protocol Corporate Standard](https://ghgprotocol.org/corporate-standard).

The following data sets are required to use ghgtools and calculate GHG emissions. When you library(ghgtools), these data sets are made available.

**Emission_Factor_Library** - The catalog of emission factors for a variety of activities. Our team consolidates emission factors from a collection of reputable sources, such as the EPA's [Emission Factor Hub](https://www.epa.gov/climateleadership/ghg-emission-factors-hub). Unless you want to add your own custom emission factors (which you certainly can, these tools are open source after all!), you don't need to worry about the emission factor library. It is pre-populated and ready to go. 

**Asset_Portfolio** - The list of all your assets. This likely includes buildings and vehicles. You may also designate an _Enterprise_ asset to calculate scope 3 emissions for purchasing, business travel, and other supply chain related activities. 

**Activity_Data** - The record of energy consumption and other GHG-producing activities across each of your assets. 

## How it Works

First, we need to load ghgtools
<br>

```{r setup}
library(ghgtools)
```

<br>
The emission factor library, `EFLibrary`, is loaded from the ghgtools package. There is a row of data for each greenhouse gas produced by the activity. For example, burning natural gas will generate CO2, CH4 and N2O. `EFLibrary` is loaded when you library ghgtools.
<br>

```{r, include=TRUE, message=FALSE, warning=FALSE}
library(kableExtra)
EFLibrary[is.na(EFLibrary)] <- ""
kbl(head(EFLibrary, 100)) %>%
  kable_material(c("striped", "hover")) %>%
  scroll_box(width = "100%", height = "400px")
```
<br>
If you want to export the EFLibrary and consolidate each emission factor into carbon dioxide equivalents (CO2e), you can use `export_efl`. When using `export_efl`, input your selection of GWPs (choices are "SAR", "AR4", "AR5", or "AR6"). If you have no preference for GWPs, we recommend following UNFCCC guidelines, which require the use of GWP values from the IPCC's Fifth Assessment Report (AR5).
<br>

```{r, include=TRUE, message=FALSE, warning=FALSE}
AR5_EFL <- export_efl("AR5")
AR5_EFL[is.na(AR5_EFL)] <- ""
kbl(head(AR5_EFL, 100)) %>%
  kable_material(c("striped", "hover")) %>%
  scroll_box(width = "100%", height = "400px")
```

<br>
Now we need to create our data input templates in the working directory. Use `create_templates()`
<br>

```{r, include=TRUE, message=FALSE, warning=FALSE}
create_templates()
```

<br>
The next steps are provided in the return message from `create_templates()`. It's time to load you data into **AssetPortfolio.csv** and **ActivityData.csv**. If you'd just like to experiment with the functionality of ghgtools, you can leave the templates as is. There is some dummy data pre-populated in the templates. Once your data is loaded, run `check_load_data()` to load data from the input templates into global environment variables.
<br>

```{r, include=TRUE, message=FALSE, warning=FALSE}
load_data()
```

<br>
Great! The data loaded successfully. Let's take a look at the emission factor library
<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
Activity_Data[is.na(Activity_Data)] <- ""
kbl(head(Activity_Data, 100)) %>%
  kable_material(c("striped", "hover")) %>%
  scroll_box(width = "100%", height = "400px")
```

<br>
And our Asset Portfolio
<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
Asset_Portfolio[is.na(Asset_Portfolio)] <- ""
kbl(head(Asset_Portfolio, 100)) %>%
  kable_material(c("striped", "hover")) %>%
  scroll_box(width = "100%", height = "400px")
```

<br>
The core function of ghgtools is `ghg_inventory()`. This function requires activity data, an asset portoflio, and a GWP selection. `ghg_inventory()` will default to the internal emission factor library `EFLibrary` if no selection is made. You may create your own EFL and use it in `ghg_inventory()` if you like. 
<br>

```{r, include=TRUE, message=FALSE, warning=FALSE}
GHG_rawdata <- ghg_inventory(Activity_Data, Asset_Portfolio, "AR5")
```
<br>
Let's take a look at the ghg inventory that was calculated
<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
kbl(head(GHG_rawdata, 100)) %>%
  kable_material(c("striped", "hover")) %>%
  scroll_box(width = "100%", height = "400px")
```

<br>
`ghg_inventory()` has merged our AssetPortoflio, ActivityData, and EFL to generate a GHG emissions inventory report. Let's take a quick look at how we might visualize this data. 
<br>

```{r, echo=FALSE, message = FALSE, warning=FALSE, fig.width=6, fig.height=4}
library(ggplot2)
ghg_summary <- GHG_rawdata[, .(EmissionTotal = sum(mt_co2e)), by = emission_scope]
ghg_summary[, percentage := (EmissionTotal / sum(EmissionTotal))*100]
GHG_sum_chart <- ggplot(ghg_summary, aes(x = emission_scope, y = EmissionTotal)) +
  geom_bar(stat = "identity", width = 0.8, fill = "grey", color = "black") +
  geom_text(aes(label = paste0(round(percentage), "%")),
            position = position_stack(vjust = 0.5),
            size = 4,
            fontface = "bold",
            color = "black") +
  labs(title = "GHG Emissions Total",
       x = "",
       y = "MT CO2e")
print(GHG_sum_chart)
```

<br>

```{r, echo=FALSE, message = FALSE, warning=FALSE, fig.width=6, fig.height=4}
Ecat_summary <- GHG_rawdata[, .(EmissionTotal = sum(mt_co2e)), by = service_type]
Ecat_sum_chart <- ggplot(Ecat_summary, aes(x = reorder(service_type, EmissionTotal), y = EmissionTotal)) +
  geom_bar(stat = "identity", width = 0.7, fill = "darkgreen", color = "black") +
  labs(title = "GHG Emissions by Activity Type",
       x = "",
       y = "MT CO2e") +
  coord_flip()
print(Ecat_sum_chart)
```

<br>

```{r, echo=FALSE, message = FALSE, warning=FALSE,fig.width=6, fig.height=4}
scope1and2 <- c("Scope 1", "Scope 2")
state_data <- GHG_rawdata[emission_scope %in% scope1and2, .(EmissionTotal = sum(mt_co2e)), by = state]
state_data[, EmissionTotal := round(EmissionTotal, 0)]
state_data[, mid := EmissionTotal - (0.5 * EmissionTotal)]
state_data_chart <- ggplot(state_data, aes(x = state, y = EmissionTotal)) +
  geom_bar(stat = "identity", width = 0.5, fill = "lightgrey", color = "black") +
   geom_text(aes(x = state, y = mid, label = paste0(EmissionTotal, " MT CO2e")), 
            size = 4,
            fontface = "bold", 
            color = "black") +
  labs(title = "Scope 1 & 2 Emissions by State",
       x = "",
       y = "") +
  coord_flip()
print(state_data_chart)
```

